Basics of Hardware Cache Optimization

### 2017-06-25 tex/com.tex
Minor tweaks and additions

\alts{When \(D^\mathsf{T}\) has, For} sufficiently long rows (\(m > B\))\footnote{%
   \(D^\mathsf{T}\) will be an \(n\times m\) matrix, so \(m\) is its number of columns.
},
every consecutive access will be to a different cache line.  If it has sufficiently many
rows, no access will be to a cache line that is still loaded; once the inner loop
completes, the cache line holding \alts[2]{\mintinline{c}{E[0][i]}, \(B_{0,i}\)} will have
been evicted.%
\footnote{%
   assuming LRU or FIFO replacement % FIXME: but we use optimal replacement!
}
Therefore, this algorithm incurs \(\Theta(mn)\) cache misses.

The algorithm is \alts{by definition, technically} already cache-oblivious since it
doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
with a divide-and-conquer approach.
% The code may only be 72 characters wide (75 when counting the indentation that will be
% stripped).  Note that the array arguments will decay to pointers.
\begin{minted}[autogobble, mathescape]{c}
   // Transpose the submatrix $(d_{ij})_{i\in I,\:j\in J}$.
   void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
      int num_rows = 1 + I[1] - I[0];
      int num_cols = 1 + J[1] - J[0];
      if (num_cols == 1 && num_rows == 1) {
         E[J[0]][I[0]] = D[I[0]][J[0]];
      } else if (num_cols <= num_rows) {
         // Horizontally slice D into two submatrices and recurse.
         transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
         transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
      } else { /* Vertically slice D analogously... */ }
   }
\end{minted}
% // Transpose the submatrix $(d_{ij}),\: i\in I, j\in J$.
% // Transpose the submatrix $(d_{ij})_{I_0<i<I_1, J_0<j<J_1}$
% // Transpose the submatrix $D\left[...\right]$

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256},
               },
              },
       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
         % ticks={major also at={1}}
       },
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=xpose/speedup.csv, separator=\space];
   \end{tikzpicture}
   \caption{TODO}
   \label{fig:TODO}
\end{figure}

>>>>

\alts{When \(D^\mathsf{T}\) has, For} sufficiently long rows (\(m > B\))\footnote{%
   \(D^\mathsf{T}\) will be an \(n\times m\) matrix, so \(m\) is its number of columns.
},
every consecutive access will be to a different cache line.  If it has sufficiently many
rows,
\alts{{%
      the constant amount of lines that are still kept in cache by the optimal replacement
      strategy once the inner loop completes and we need them again becomes
      negligible.\x{\footnote{%
         With LRU or FIFO replacement, ...
   }}}, {
      no access will be to a cache line that is still loaded; once the inner loop
      completes, the cache line holding \alts[2]{\mintinline{c}{E[0][i]}, \(B_{0,i}\)}
      will have been evicted.%
      \footnote{%
         assuming LRU or FIFO replacement % XXX: but we use optimal replacement!
      }%
   },
}
Therefore, this algorithm incurs \(\Theta(mn)\) cache misses.

The algorithm is \alts{by definition, technically} already cache-oblivious since it
doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
with a divide-and-conquer approach.  The idea is to recursively divide the input matrix
\(D\) into two equal-sized submatrices along the greater dimension.  If \(m \geq n\) (more
rows than columns), let
\begin{equation*}
   D = \begin{bmatrix}
      D_1\\
      D_2
   \end{bmatrix}
\end{equation*}
and use \(D^\mathsf{T} = \begin{bmatrix} D_1^\mathsf{T} & D_2^\mathsf{T} \end{bmatrix}\)
to compute the transpose.  If \(m < n\), analogously split \(D\) vertically.  Eventually,
the submatrices will become small enough so that pairs of input and output submatrices fit
into cache at the same time, at which point it doesn't matter in what order we access the
elements.\footnote{%
   No cache line will have to be evicted once loaded.
}
The recursion continues all the way down to \(1 \times 1\) submatrices, but this doesn't
change the theoretical analysis.  According to
\textcite[theorem 2 and 3, \ppno~19--21]{coa-thesis}
% \textcite[theorem 2 and 3, \ppno~19\psqq]{coa-thesis}
this algorithm incurs an optimal amount of \(\Theta(1 + mn / B)\) cache misses.
% This reduces the amount of cache misses to \(\Theta(1 + \nicefrac{mn}{B})\) according to
% \textcite[theorem 2, \ppno~19\psq]{coa-thesis}.

In practice, the new algorithm performs as badly as one might expect from this excessive
use of recursion.  \Cref{fig:xpose-speedup} gives the speedup compared to the
straightforward implementation, which is usually faster.

% The code may only be 72 characters wide (75 when counting the indentation that will be
% stripped).  Note that the array arguments will decay to pointers.
\begin{minted}[autogobble, mathescape]{c}
   // Transpose the submatrix $(d_{ij})_{i\in I,\:j\in J}$.
   void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
      int num_rows = 1 + I[1] - I[0];
      int num_cols = 1 + J[1] - J[0];
      if (num_cols == 1 && num_rows == 1) {
         E[J[0]][I[0]] = D[I[0]][J[0]];
      } else if (num_cols <= num_rows) {
         // Horizontally slice D into two submatrices and recurse.
         transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
         transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
      } else { /* Vertically slice D analogously... */ }
   }
\end{minted}
% // Transpose the submatrix $(d_{ij}),\: i\in I, j\in J$.
% // Transpose the submatrix $(d_{ij})_{I_0<i<I_1, J_0<j<J_1}$
% // Transpose the submatrix $D\left[...\right]$

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256},
               },
              },
       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
         % ticks={major also at={1}}
       },
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=xpose/speedup.csv, separator=\space];
   \end{tikzpicture}
   \caption{TODO}
   \label{fig:xpose-speedup}
\end{figure}
